\doctitle

### 2023-05-23 cs.tex
Move current results on CS to the thesis

% Recent progress on deep SPI
Because each CS measurement contains highly-compressed information about the observed frame, one needs to sample a large number of such measurements. Thus, many recent works propose approaches to reduce the number of required samples per frame (SPF). These approaches can be roughly grouped into two buckets: signal-dependent control of the mirrors and more sophisticated reconstruction techniques. The former ensures that each single-pixel measurement contains as much information as possible by focusing the mirrors on the selected regions of the frame~\cite{zhang2017fast,sun2017russian,xu20181000}. Although highly effective, such an approach cannot be applied to detectors that do not offer fine-grained control of their mirrors. The latter group focuses on creating more efficient optimization algorithms for the decoding step~\cite{katz2009compressive}. It includes non-iterative algorithms such as differential ghost imaging (DGI) \cite{gong2010method}, as well as iterative methods such as total variation (TV)~\cite{suo2016signal} regularization. Notably, recent approaches adopted deep learning (DL) for SPI reconstruction. These approaches showed that using deep neural networks can dramatically reduce sampling ratio and offer near-real-time performance \cite{lyu2017deep,higham2018deep,wang2019learning,wang2022single}. However, most of these methods use multiple samples for single-frame reconstruction and do not yield satisfactory reconstruction of multiple sequential frames when the number of samples per frame is small, even if the total number of samples is large, as we show below.

@@

\paragraph{Reduced-Order Model with Non-Linear Latent Dynamics} We assume the same notation as in Section~\ref{sec:method} and define a reduced-order model (ROM) as a tuple $(\psi_\theta, \phi_\theta, h_\theta)$, which consists of a decoder $\psi_\theta(z)$, an encoder $\phi_\theta(z)$, and the latent dynamics $h(z)$, respectively, all parametrized by trainable weights $\theta$. We also assume that it is technically possible\footnote{For example, if $h(z)$ is a neural network, then one can evaluate such an integral using adjoint-sensitivity-based techniques, e.g. \texttt{torchdiffeq} package by \cite{chen2018neuralode}.} to evaluate a differentiable integral over the latent dynamics $h(z)$ for a given period of time. Figure~\ref{fig:cs_abstract} illustrates the architecture of the ROM. Since this section assumes that such a ROM comes pre-trained, we consider the weights $\theta$ to be frozen and omit them from the notation.

@@

In particular, we consider a single pixel camera setup in which, for every time instance $t$, $p$ acquisitions $y(t)$ are obtained by a high sampling rate photo-detector using the projection matrix $A_t$. The rows of the matrix $A_t$ correspond to a binary mask pattern that can be encoded using a digital micro-mirror device (DMD, \cite{pittman1995optical, sampsell1993overview}). Similarly to the setup from~\cite{duarte2008single}, the mirrors of our (idealized) DMD array represent pixels of the future reconstructed image, and the slopes of the mirrors represent the weights. The light from a scene lands on a DMD, which slopes each of its mirrors to reflect light either towards the detector (+1) or away from it (-1) according to a uniform Bernoulli variable. Next, the (+1) pixels go through a double-convex lens that focuses them on a signal photon detector. Finally, the detector integrates the signal and provides the final observation as its output voltage, which is later digitized by an A/D converter. The slopes of each mirror are also recorded in the detector's memory. Figure~\ref{fig:SPI} illustrates an example of the single pixel imaging setup where a gas plume is imaged using a DMD array and a medium infrared (MIR) photo-detector.

A modern DMD supports snapshot frequencies of around 10~kHz (\cite{wang2022single}). When the dynamics of the recorded scene are much slower, one can assume that such a device can take multiple measurements simultaneously\footnote{When the dynamics change rapidly, one would have to use an array of DMD devices to achieve more than one sample per frame}. The number of samples recorded per  frame is referred to as \textit{samples per frame}, or SPF. It is often divided by the total number of mirrors of the DMD device and multiplied by 100\%, in which case it is called a \textit{sample per frame rate} $\beta$. For a compressive-sensing-based approach to be practically useful it should require a relatively small $\beta$ (e.g. 2-10\%) for a successful reconstruction  of the scene.

@@

We compare our approach (PINODE) with three alternatives: Digital Ghost Imaging by~\cite{ferri2010differential} and~\cite{gong2010method}, Total Variation Regularization  (TVR) by~\cite{bian2018experimental}, and an approach by~\cite{bora2017compressed} which uses an autoencoder (AE). We measure reconstruction accuracy with three commonly-used metrics: the residual mean-squared error (RMSE), the peak signal-to-noise ratio (PSNR), and the structural similarity index measure (SSIM). We evaluate both metrics for Burger's trajectories as 2D-images in spatial and in temporal domains.

@@

\begin{figure}
    \centering
    	\includegraphics[width=\textwidth]{figures/cs_burgers_comparison_64.pdf}
    \caption{\label{fig:cs_burgers_example_large_spf} Example reconstruction with 64 samples per frame (SPF). Unsurprisingly, all handle the job well since SPF is equal to the number of pixels.}
\end{figure}

\begin{figure}
    \centering
    	\includegraphics[width=\textwidth]{figures/cs_burgers_comparison_4.pdf}
    \caption{\label{fig:cs_burgers_example_small_spf}Example reconstruction with 4 samples per frame (SPF). All algorithms except our proposed method fail to recover the image due to the insufficient SPF rate (~6\%). In contrast, our proposed method achieves nearly the same reconstruction quality as in Figure~\ref{fig:cs_burgers_example_large_spf} with SPF rate of 100\%.}
\end{figure}

\paragraph{Results} The results are presented in Figure~\ref{fig:cs_burgers_spf}. We plot median PSNR, RMSE, and SSIM over different number of samples per frame taken. We see that PINODE achieves 70\% higher peak signal-to-noise-ratio (PSNR) and 4 times higher structural similarity index measure (SSIM) when SPF is low relative to its nearest competitor (AE). We also see that the DGI and TVR approaches do not reconstruct images faithfully until SPF approaches the number of pixels in a snapshot; Figure~\ref{fig:cs_burgers_example_large_spf} provides an example of high-SPF reconstruction. In contrast, PINODE is able to reconstruct most trajectories with high accuracy using a very small number of samples per frame e.g. 4, as visualized on Figure~\ref{fig:cs_burgers_example_small_spf}. This happens because the reduced-order model of PINODE serves as a strong prior: the model reconstructs the trajectory as a whole instead of reconstructing every snapshot separately, as all other algorithms do in this comparison. It allows PINODE to borrow strength across snapshots, even when SPF per each snapshot is small.

\subsubsection{Kolmogorov Flow}
We study the performance of our framework on Kolmogorov-like flow first-described by~\cite{dovzhenko1981generation} and later analyzed by \cite{tithof2017bifurcations,suri2017forecasting}. The following equation describes the behavior of a 2D velocity field $\bd{u}(x, y, t)$:

@@

To train a ROM which predicts $u(x, y, t)$ we generate 8192 data trajectories as solutions of~\eqref{eq:kolmogorov_flow_definition} using a spectral solver\footnote{\href{https://github.com/zhong1wan/data-assisted/blob/master/Kolmogorov/kol2d_odd.py}{https://github.com/zhong1wan/data-assisted/blob/master/Kolmogorov/kol2d\_odd.py}} from~\cite{wan2018data}. Each solution was simulated from t=0 to t=100 to get past the transient regime and then recorded from t=100 to t=110, with the time-step $dt=0.5$. As in Section~\ref{sec:method}, the ROM consists of an encoder $\phi(u)$, a decoder $\psi(z)$, and the latent dynamics $h(z)$. It was trained by minimizing the data-driven loss~\eqref{eq:loss_data_driven} until the prediction RMSE on a holdout set stopped improving.

@@

A sample result is displayed on Figure~\ref{fig:kolmogorov_example}. The rows represent three different moments in time. The first column represents the true flow. The second column represents the prediction of the ROM if the initial state of the system were to be known. It is unknown, so the second row serves as a reference for the maximal theoretical performance of a ROM. The third column displays the results of CS reconstruction of our algorithm with 512 samples per frame. Each frame is $66\times 66=4356$ pixels, which yields a sampling-per-frame rate of 12\%. We observe that the reconstruction faithfully recovered the signal. Moreover, we see that the CS algorithm is not limited in its performance by the performance of the underlying ROM: the third row indicates that the model is unable to completely recover shifting dynamics, whereas the CS algorithm aids the model's predictions with CS observations and fully reconstructs the signal.

>>>>

% Recent progress on deep SPI
Because each CS measurement contains highly-compressed information about the observed frame, one needs to sample a large number of such measurements. Thus, many recent works propose approaches to reduce the number of required samples per frame (SPF). These approaches can be roughly grouped into two buckets: signal-dependent control of the mirrors and more sophisticated reconstruction techniques. The former ensures that each single-pixel measurement contains as much information as possible by focusing the mirrors on the selected regions of the frame~\cite{zhang2017fast,sun2017russian,xu20181000}. Although highly effective, such an approach cannot be applied to monitoring scenes that don't have a ``main zone'' where the information acquisition could be focused. For example, a camera monitoring a gas tank for leaks can not have a mask that works well for leaks appearing only in the middle of the scene. The second group of approaches focuses on creating more efficient optimization algorithms for the decoding step~\cite{katz2009compressive}. It includes non-iterative algorithms such as differential ghost imaging (DGI) \cite{gong2010method}, as well as iterative methods such as total variation (TV)~\cite{suo2016signal} regularization. Notably, recent approaches adopted deep learning (DL) for SPI reconstruction. These approaches showed that using deep neural networks can dramatically reduce sampling ratio and offer near-real-time performance \cite{lyu2017deep,higham2018deep,wang2019learning,wang2022single}. However, most of these methods use multiple samples for single-frame reconstruction and do not yield satisfactory reconstruction of multiple sequential frames when the number of samples per frame is small, even if the total number of samples is large, as we show below.

@@

\paragraph{Reduced-Order Model with Non-Linear Latent Dynamics} We assume the same notation as in Section~\ref{sec:method} and define a reduced-order model (ROM) as a tuple $(\psi_\theta, \phi_\theta, h_\theta)$, which consists of a decoder $\psi_\theta(z)$, an encoder $\phi_\theta(z)$, and the latent dynamics $h(z)$, respectively, all parametrized by trainable weights $\theta$. We also assume that it is technically possible\footnote{For example, if $h(z)$ is a neural network, then one can evaluate such an integral using adjoint-sensitivity-based techniques, e.g. \texttt{torchdiffeq} package by \cite{chen2018neuralode}.} to evaluate a differentiable integral over the latent dynamics $h(z)$ for a given period of time. Figure~\ref{fig:cs_abstract} illustrates the architecture of the ROM. We assume that such ROM comes fully trained, so we consider its weights $\theta$ to be frozen and omit them from the notation.

@@

We consider a single pixel camera setup in which, for every time instance $t$, $p$ acquisitions $y(t)$ are obtained by a high sampling rate photo-detector using the projection matrix $A_t$. The rows of the matrix $A_t$ correspond to a binary mask pattern that can be encoded using a digital micro-mirror device (DMD, \cite{pittman1995optical, sampsell1993overview}). Similarly to the setup from~\cite{duarte2008single}, the mirrors of our (idealized) DMD array represent pixels of the future reconstructed image, and the slopes of the mirrors represent the weights. The light from a scene lands on a DMD, which slopes each of its mirrors to reflect light either towards the detector (+1) or away from it (-1) according to a uniform Bernoulli variable. Next, the (+1) pixels go through a double-convex lens that focuses them on a signal photon detector. Finally, the detector integrates the signal and provides the final observation as its output voltage, which is later digitized by an A/D converter. The slopes of each mirror are also recorded in the detector's memory. Figure~\ref{fig:SPI} illustrates an example of the single pixel imaging setup where a gas plume is imaged using a DMD array and a medium infrared (MIR) photo-detector.

A modern DMD supports snapshot frequencies of around 10~kHz (\cite{wang2022single}). When the scene changes much slower one can assume that such a device can take multiple measurements of a scene simultaneously\footnote{When the dynamics change rapidly, one would have to use an array of DMD devices to achieve more than one sample per frame}. The number of samples recorded per  frame is referred to as \textit{samples per frame}, or SPF. It is often divided by the total number of mirrors of the DMD device and multiplied by 100\%, in which case it is called a \textit{sample per frame rate} $\beta$. For a compressive-sensing-based image reconstruction approach to be practically useful it should require a relatively small rate $\beta$ (e.g. 2-10\%) for a successful reconstruction  of the scene.

@@

We compare our approach (PINODE) with four alternatives: Digital Ghost Imaging (DGI) by~\cite{ferri2010differential} and~\cite{gong2010method}, its Autoencoder-enhanced variation (DGI+Conv. Decoder) by~\cite{wang2022single}, Total Variation Regularization  (TVR) by~\cite{bian2018experimental}, and an approach by~\cite{bora2017compressed} which uses an autoencoder (AE). We measure reconstruction accuracy with three commonly-used metrics: the residual mean-squared error (RMSE), the peak signal-to-noise ratio (PSNR), and the structural similarity index measure (SSIM). We evaluate both metrics for Burger's trajectories as 2D-images in spatial and in temporal domains.

@@

\begin{figure}
    \centering
    	\includegraphics[width=\textwidth]{figures/cs_burgers_comparison_64.pdf}
    \caption{\label{fig:cs_burgers_example_large_spf} Example reconstruction with 32 samples per frame (SPF), an SPF rate of 50\%. Rather unsurprisingly, all algorithms complete the reconstruction faithfully due to a large amount of samples.}
\end{figure}

\begin{figure}
    \centering
    	\includegraphics[width=\textwidth]{figures/cs_burgers_comparison_4.pdf}
    \caption{\label{fig:cs_burgers_example_small_spf}Example reconstruction with 2 samples per frame (SPF). All algorithms except our proposed method fail to recover the image due to the insufficient SPF rate (~3\%). In contrast, our proposed method achieves nearly the same reconstruction quality as in Figure~\ref{fig:cs_burgers_example_large_spf} with SPF rate of 50\%.}
\end{figure}

\paragraph{Results} The results are presented in Figure~\ref{fig:cs_burgers_spf}. We plot median PSNR, RMSE, and SSIM over different number of samples per frame taken. We see that PINODE achieves 70\% higher peak signal-to-noise-ratio (PSNR) and 4 times higher structural similarity index measure (SSIM)  relative other algorithms when SPF is low. We also see that the DGI and TVR approaches do not reconstruct images faithfully until SPF approaches the number of pixels in a snapshot; Figure~\ref{fig:cs_burgers_example_large_spf} provides an example of high-SPF reconstruction. In contrast, PINODE is able to reconstruct most trajectories with high accuracy using a very small number of samples per frame e.g. two samples, as visualized on Figure~\ref{fig:cs_burgers_example_small_spf}. It happens because the reduced-order model of PINODE serves as a strong prior: the model reconstructs the trajectory as a whole instead of reconstructing every snapshot separately, as all other algorithms do in this comparison. It allows PINODE to borrow strength across snapshots, even when SPF per each snapshot is small.

We also study robustness of PINODE approach to noise in compressive samples $y$ relatively to other algorithms. Figure~\ref{fig:cs_burgers_noise_aggregate} presents an aggregated comparison of reconstruction of solutions of Burger's equations under noise for SPF=12.5\%; the setup is equivalent to the one for SPF = 8 on Figure~\ref{fig:cs_burgers_spf} . Namely, we plot performance metrics (PSNR, RMSE, SSIM) of reconstructions from different models against Signal-to-Noise Ratio (SNR) in the vector of compressive sensing single-pixel observations $y$. The performance of PINODE remains superior until all models start reconstructing the scene very poorly. Figure~\ref{fig:cs_burgers_noise_examples} visualizes reconstructions for selected levels of SNR. We see that PINODE is able to provide a faithful reconstruction even under a substantial noise (e.g. third column).

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figures/cs_burgers_noise_aggregate.png}
	\caption{\label{fig:cs_burgers_noise_aggregate} Performance of reconstruction of solutions of Burger's equations under noise for SPF=12.5\%. We plot the performance metrics of different models against Signal-to-Noise Ratio (SNR) in the vector of compressive sensing single-pixel observations $y$. The performance of PINODE remains superior until all models start reconstructing the scene very poorly. See Figure~\ref{fig:cs_burgers_noise_examples} for example visualizations.}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figures/cs_burgers_noise_examples.pdf}
	\caption{\label{fig:cs_burgers_noise_examples} Examples of compressive-sensing reconstructions of trajectories of Burger's equation under noise from Figure~\ref{fig:cs_burgers_noise_aggregate}.}
\end{figure}

\subsubsection{Kolmogorov Flow}
\label{sec:kolmogorov_flow}
We study the performance of our framework on Kolmogorov-like flow first-described by~\cite{dovzhenko1981generation} and later analyzed by \cite{tithof2017bifurcations,suri2017forecasting}. The following equation describes the behavior of a 2D velocity field $\bd{u}(x, y, t)$:

@@

To train a ROM which predicts $u(x, y, t)$ we generate 8192 data trajectories as solutions of~\eqref{eq:kolmogorov_flow_definition} using a spectral solver\footnote{\href{https://github.com/zhong1wan/data-assisted/blob/master/Kolmogorov/kol2d_odd.py}{https://github.com/zhong1wan/data-assisted/blob/master/Kolmogorov/kol2d\_odd.py}} from~\cite{wan2018data}. Each solution was simulated from t=0 to t=100 to get past the transient regime and then recorded from t=100 to t=110, with the time-step $dt=0.5$. As in Section~\ref{sec:method}, the ROM consists of an encoder $\phi(u)$, a decoder $\psi(z)$, and the latent dynamics $h(z)$. It was trained by minimizing the data-driven loss~\eqref{eq:loss_data_driven} until the prediction RMSE on a holdout set stopped improving, which took about 60 epochs.

@@

A sample result is displayed on Figure~\ref{fig:kolmogorov_example}. The rows represent three different moments in time. The first column represents the true flow. The second column represents the prediction of the ROM if the initial state of the system were to be known to the model. It is unknown, so the second row serves as a reference performance of an underlying ROM. The third column displays the results of CS reconstruction of our algorithm with 512 samples per frame. Each frame is $66\times 66=4356$ pixels, which yields a sampling-per-frame rate of 12\%. We observe that the reconstruction faithfully recovered the signal. Moreover, we see that the CS algorithm is not limited in its performance by the performance of the underlying ROM: the third row indicates that the model is unable to completely recover shifting dynamics, whereas the CS algorithm aids the model's predictions with CS observations and fully reconstructs the signal.

\subsubsection{Methane Leaks Data}
In this section we use our technique to reconstruct videos of gas leaks using observations from a single-pixel camera. We use a subset of GasVid dataset by~\cite{wang2020machine} which consists of recordings of a methane gas leaks. 10 videos 10 to 20 seconds each. We split them into 1-second-long intervals, where each interval contains 10 steps: $T = 1$ sec., $dt = 0.1$ sec. Next, we remove backgrounds from each video using Gaussian Mixture-based Background/Foreground Segmentation (MOG2) algorithm by~\cite{zivkovic2006efficient,zivkovic2004improved} implemented in OpenCV library~(\cite{opencv_library}). Finally, we split all intervals into four non-overlapping batches: train (390), dev (50), test (49), and compressive sensing (2). We use the first three to train, fine-tune, and select a final ROM respectively, and then we use the fourth batch for compressive sensing reconstruction. We use the same architecture of ROM as in Section~\ref{sec:kolmogorov_flow}. At the compressive sensing step we sample 512 single-pixel observations per each $240 \times 320$ video frame; it corresponds to SPF ratio of $0.66\%$ (2/3 of one percent).

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figures/cs_gas_examples.pdf}
	\caption{\label{fig:cs_gas_examples}}
\end{figure}

The example reconstructions are presented on Figure~\ref{fig:cs_gas_examples}. We see that the algorithm was able to faithfully reconstruct the trajectories. Although it was unable to capture the details of turbulent flows, it was able to predict the altitude, the direction, and the spread of the plumes correctly, which are the most important aspects for practical applications in real-world gas monitoring applications.

We note the threefold difficulty of the task. First, the background subtraction struggles to separate grey smoke from a dynamic grey background, which leads to fantom clouds of smoke in processed data. Second, the amount of training data (3900 frames) is extremely small for training a large convolutional ROM. For example, \cite{wang2020machine} used the same dataset for training a simpler classification network (leak/non-leak) using ~11000 frames. Finally, an SPF of 0.66\% is vastly insufficient for frame-by-frame reconstruction, as most state-of-the-art methods require ~5-10\% SPF, depending on the nature of the picture (	\cite{wang2022single}). Thus, the inductive bias of the ROM becomes a crucial component for a successful video reconstruction.
